The Spatial and Temporal Domains of Modern Ecology

### 2017-05-15 Rmd/space-time.Rmd
Added Figure 3 to text and associated text.  Edited Figure 3 to add percetage values for each bin atop whiskers.

# image(rres3, col = rev(terrain.colors(100)))
# image(rext3, col = rev(terrain.colors(100)))
# datf[, points(log10(eff_dur), log10(eff_ext), pch = 20, cex = 0.5)]
# datf[like(study_type, "remote"), {
#   points(log10(eff_dur), log10(eff_ext), pch = 20, cex = 0.5, col = "blue")
# }]
# points(resdat3$x, resdat3$y, pch = 20, cex = 0.5)
# image(sres, col = rev(terrain.colors(100)))
# datf[, points(log10(plot_res), log10(eff_ext), pch = 20, cex = 0.5)]
# datf[like(study_type, "remote"), {
#   points(log10(plot_res), log10(eff_ext), pch = 20, cex = 0.5, col = "blue")
# }]
# datf[like(study_type, "field"), {
#   points(log10(plot_res), log10(eff_ext), pch = 20, cex = 0.5, col = "red")
#   abline(lm(log10(plot_res) ~ log10(eff_ext)))
# }]
# points(spatdat$x, spatdat$y, pch = 20, cex = 0.5)
```

@@

# calculate box statistics
# Extent differences
eint <- aaxis2$logarea
edif_int <- rbindlist(lapply(hdat_bs2, function(y) {  # y <- hdat_bs2[[2]]
  dd <- copy(y$edb[!is.na(aext)])
  dd[aext < ex[1], aext := ex[1]]  # set to min
  # dd[, list("int" = findInterval(aext, eint), "dif" = dif)]
  dd[, int := findInterval(aext, eint, rightmost.closed = TRUE)]
  # dd[, list("dif" = mean(dif, na.rm = TRUE), "N" = .N), by = int][order(int)]
}))
ints <- sort(edif_int[, unique(int)])
ebox <- lapply(ints, function(x) edif_int[int == x, lmisc::box_stats(dif)])
pctobs <- sapply(ints, function(x) edif_int[int == x, .N])
ewgt <- pctobs / sum(pctobs)
# ealphs <- (ewgt - min(ewgt)) / diff(range(ewgt))
ewgtint <- c(0, 1, 5, 10, 15, 20, 25) / 100  # weight intervals
ewgttxt <- c("  0", "  1", "  5", seq(10, 25, 5))
ecols <- findInterval(ewgt, ewgtint)  # corresponding color index

# temporal differences
dint <- taxis2$logdays[-5]
tx <- range(dint)
ddif_int <- rbindlist(lapply(hdat_bs2, function(y) {  # y <- hdat_bs2[[2]]
  dd <- copy(y$tdb[!is.na(adur)])
  dd[adur < tx[1], adur := tx[1]]  # set to min
  dd[adur > tx[2], adur := tx[2]]  # set to max
  dd[, int := findInterval(adur, dint, rightmost.closed = TRUE)]  # include high
}))
dints <- sort(ddif_int[, unique(int)])
tbox <- lapply(dints, function(x) ddif_int[int == x, lmisc::box_stats(dif)])
pctobs <- sapply(dints, function(x) ddif_int[int == x, .N])
dwgt <- pctobs / sum(pctobs)
# dalphs <- (dwgt - min(dwgt)) / diff(range(dwgt))
dwgtint <- c(0, 1, 5, 10, 15, 20, 25, 30) / 100  # weight intervals
dwgttxt <- c("  0", "  1", "  5", seq(10, 30, 5))
dcols <- findInterval(dwgt, dwgtint)

@@

# pdf("paper/figures/act_v_eff_diff.pdf", width = 7, height = 4)
png("paper/figures/act_v_eff_diff.png", width = 7, height = 4, res = 600, 
    units = "in")
par(mfrow = c(1, 2), mar = c(4, 0, 1, 0.75), oma = c(0, 3, 0, 0))
bx <- rowMeans(cbind(eint[-length(eint)], eint[-1]))
# cols <- sapply(ealphs, function(x) rgb(0, 0.1, 1, alpha = x))
cols <- brewer.pal(length(ewgtint) - 1, "Blues")
plot(ex, c(0, 14), pch = "", xlab = "", ylab = "", xaxt = "n", yaxt = "n",
     mgp = c(3, 0.5, 0), tcl = -0.2, xaxs = "i")
polygon(c(ex, rev(ex), ex[1]), c(-1, -1, 15, 15, -1), col = "grey90")
ylines(yr = c(0, 14), by = 2, col = "grey95")
for(i in 1:length(ebox)) {
  boxplot_v(bx[i], y = ebox[[i]], n = 1.75, whiskhgt = 1.5, 
            bfill = cols[ecols][i])
} 
axis(1, at = eint, labels = alab2, las = 2, tcl = -0.2, cex.axis = cxa, 
     mgp = c(3, 0.5, 0))
axis(2, at = seq(0, 14, 2), las = 2, tcl = -0.2, cex.axis = cxa, 
     mgp = c(3, 0.5, 0))
mtext("Actual Extent", side = 1, line = 3, cex = 0.9)
mtext("Difference Between", side = 2, line = 2.1, cex = 0.9)
mtext("Effective and Actual (Decades)", side = 2, line = 1.4, cex = 0.9)
lrng <- c(mean(ex), max(ex) - 0.05 * max(ex))
xs <- seq(lrng[1], lrng[2], by = (diff(lrng)) / (length(cols)))
rect_shade(xs, y = c(11.5, 12.5), fillcol = cols, linecol = "grey30")
text(xs, rep(10.5, length(xs)), ewgttxt, srt = 90, col = "grey30")
text(mean(xs), 13, "% of observations", col = "grey30")
mtext("A", side = 3, line = -1, cex = 1, adj = 0.05)

# duration
bx <- rowMeans(cbind(dint[-length(dint)], dint[-1]))
bd <- abs(dint[-length(dint)] - dint[-1])
# cols <- sapply(dalphs, function(x) rgb(1, 0, 0.1, alpha = x))
cols <- brewer.pal(length(dwgtint) - 1, "Reds")
plot(tx, c(0, 14), pch = "", xlab = "", ylab = "", xaxt = "n", xaxs = "i",
     mgp = c(3, 0.5, 0), tcl = -0.2, yaxt = "n")
polygon(c(tx, rev(tx), tx[1]), c(-1, -1, 15, 15, -1), col = "grey90")
ylines(yr = c(0, 14), by = 2, col = "grey95")
for(i in 1:length(tbox)) {
  boxplot_v(bx[i], y = tbox[[i]], n = 0.9, inhgt = bd[i] * 10, whiskhgt = 0.5, 
            bfill = cols[dcols][i])
} 
axis(1, at = dint, labels = tlab2[-c(5)], las = 2, tcl = -0.2, cex.axis = cxa, 
     mgp = c(3, 0.5, 0))
axis(2, at = seq(0, 14, 2), labels = rep("", 8), las = 2, tcl = -0.2, 
     cex.axis = cxa, mgp = c(3, 0.5, 0))
mtext("Actual Duration", side = 1, line = 3, cex = 0.9)
lrng <- c(mean(tx), max(tx) - 0.05 * max(tx))
xs <- seq(lrng[1], lrng[2], by = (diff(lrng)) / (length(cols)))
rect_shade(xs, y = c(11.5, 12.5), fillcol = cols, linecol = "grey30")
text(xs, rep(10.5, length(xs)), dwgttxt, srt = 90, col = "grey30")
text(mean(xs), 13, "% of observations", col = "grey30")
mtext("B", side = 3, line = -1, cex = 1, adj = 0.05)
dev.off()

>>>>

# image(rres3, col = rev(terrain.colors(100)))
# image(rext3, col = rev(terrain.colors(100)))
# datf[, points(log10(eff_dur), log10(eff_ext), pch = 20, cex = 0.5)]
# datf[like(study_type, "remote"), {
#   points(log10(eff_dur), log10(eff_ext), pch = 20, cex = 0.5, col = "blue")
# }]
# points(resdat3$x, resdat3$y, pch = 20, cex = 0.5)
# image(sres, col = rev(terrain.colors(100)))
# datf[, points(log10(plot_res), log10(eff_ext), pch = 20, cex = 0.5)]
# datf[like(study_type, "remote"), {
#   points(log10(plot_res), log10(eff_ext), pch = 20, cex = 0.5, col = "blue")
#   abline(lm(log10(eff_ext) ~ log10(plot_res)))
# }]
# datf[like(study_type, "field"), {
#   points(log10(plot_res), log10(eff_ext), pch = 20, cex = 0.5, col = "red")
#   abline(lm(log10(eff_ext) ~ log10(plot_res)))
# }]
# points(spatdat$x, spatdat$y, pch = 20, cex = 0.5)
```

@@

# calculate box statistics
# Extent differences
eint <- aaxis2$logarea
edif_int <- rbindlist(lapply(hdat_bs2, function(y) {  # y <- hdat_bs2[[2]]
  dd <- copy(y$edb[!is.na(aext)])
  dd[aext < ex[1], aext := ex[1]]  # set to min
  # dd[eext < ex[1], eext := ex[1]]  # set to min
  dd[, int := findInterval(aext, eint, rightmost.closed = TRUE)]
  # dd[, int := findInterval(eext, eint, rightmost.closed = TRUE)]
  # dd[, list("dif" = mean(dif, na.rm = TRUE), "N" = .N), by = int][order(int)]
}))
ints <- sort(edif_int[, unique(int)])
ebox <- lapply(ints, function(x) edif_int[int == x, lmisc::box_stats(dif)])
pctobs <- sapply(ints, function(x) edif_int[int == x, .N])
ewgt <- pctobs / sum(pctobs)
# round(cumsum(ewgt), 2)
ewgtint <- c(0, 1, 5, 10, 15, 20, 25) / 100  # weight intervals
ewgttxt <- c("  0", "  1", "  5", seq(10, 25, 5))
ecols <- findInterval(ewgt, ewgtint)  # corresponding color index

# temporal differences
dint <- taxis2$logdays[-5]
tx <- range(dint)
ddif_int <- rbindlist(lapply(hdat_bs2, function(y) {  # y <- hdat_bs2[[2]]
  dd <- copy(y$tdb[!is.na(adur)])
  dd[adur < tx[1], adur := tx[1]]  # set to min
  dd[adur > tx[2], adur := tx[2]]  # set to max
  dd[, int := findInterval(adur, dint, rightmost.closed = TRUE)]  # include high
}))
dints <- sort(ddif_int[, unique(int)])
tbox <- lapply(dints, function(x) ddif_int[int == x, lmisc::box_stats(dif)])
pctobs <- sapply(dints, function(x) ddif_int[int == x, .N])
dwgt <- pctobs / sum(pctobs)
# round(cumsum(dwgt), 2)

dwgtint <- c(0, 1, 5, 10, 15, 20, 25, 30) / 100  # weight intervals
dwgttxt <- c("  0", "  1", "  5", seq(10, 30, 5))
dcols <- findInterval(dwgt, dwgtint)

@@

# pdf("paper/figures/act_v_eff_diff.pdf", width = 7, height = 4)
png("paper/figures/act_v_eff_diff.png", width = 7, height = 4, res = 600, 
    units = "in")
par(mfrow = c(1, 2), mar = c(4, 0, 1, 0.75), oma = c(0, 3, 0, 0))
bx <- rowMeans(cbind(eint[-length(eint)], eint[-1]))
# cols <- sapply(ealphs, function(x) rgb(0, 0.1, 1, alpha = x))
cols <- brewer.pal(length(ewgtint) - 1, "Blues")
plot(ex, c(0, 14), pch = "", xlab = "", ylab = "", xaxt = "n", yaxt = "n",
     mgp = c(3, 0.5, 0), tcl = -0.2, xaxs = "i")
polygon(c(ex, rev(ex), ex[1]), c(-1, -1, 15, 15, -1), col = "grey90")
ylines(yr = c(0, 14), by = 2, col = "grey95")
for(i in 1:length(ebox)) {
  boxplot_v(bx[i], y = ebox[[i]], n = 1.75, whiskhgt = 1.5, 
            bfill = cols[ecols][i])
  text(bx[i], ebox[[i]][5] + 0.15, round(ewgt[i] * 100, 1), cex = 0.3)
} 
axis(1, at = eint, labels = alab2, las = 2, tcl = -0.2, cex.axis = cxa, 
     mgp = c(3, 0.5, 0))
axis(2, at = seq(0, 14, 2), las = 2, tcl = -0.2, cex.axis = cxa, 
     mgp = c(3, 0.5, 0))
mtext("Actual Extent", side = 1, line = 3, cex = 0.9)
mtext("Difference between Extent/Duration", side = 2, line = 2.2, cex = 0.9)
mtext("and Actual Extent/Duration (Decades)", side = 2, line = 1.4, cex = 0.9)
lrng <- c(mean(ex), max(ex) - 0.05 * max(ex))
xs <- seq(lrng[1], lrng[2], by = (diff(lrng)) / (length(cols)))
rect_shade(xs, y = c(11.5, 12.5), fillcol = cols, linecol = "grey30")
text(xs, rep(10.5, length(xs)), ewgttxt, srt = 90, col = "grey30")
text(mean(xs), 13, "% of observations", col = "grey30")
mtext("A", side = 3, line = -1, cex = 1, adj = 0.05)

# duration
bx <- rowMeans(cbind(dint[-length(dint)], dint[-1]))
bd <- abs(dint[-length(dint)] - dint[-1])
cols <- brewer.pal(length(dwgtint) - 1, "Reds")
plot(tx, c(0, 14), pch = "", xlab = "", ylab = "", xaxt = "n", xaxs = "i",
     mgp = c(3, 0.5, 0), tcl = -0.2, yaxt = "n")
polygon(c(tx, rev(tx), tx[1]), c(-1, -1, 15, 15, -1), col = "grey90")
ylines(yr = c(0, 14), by = 2, col = "grey95")
for(i in 1:length(tbox)) {
  boxplot_v(bx[i], y = tbox[[i]], n = 0.9, inhgt = bd[i] * 10, whiskhgt = 0.75, 
            bfill = cols[dcols][i])
  text(bx[i], tbox[[i]][5] + 0.15, round(dwgt[i] * 100, 1), cex = 0.3)

} 
axis(1, at = dint, labels = tlab2[-c(5)], las = 2, tcl = -0.2, cex.axis = cxa, 
     mgp = c(3, 0.5, 0))
axis(2, at = seq(0, 14, 2), labels = rep("", 8), las = 2, tcl = -0.2, 
     cex.axis = cxa, mgp = c(3, 0.5, 0))
mtext("Actual Duration", side = 1, line = 3, cex = 0.9)
lrng <- c(mean(tx), max(tx) - 0.05 * max(tx))
xs <- seq(lrng[1], lrng[2], by = (diff(lrng)) / (length(cols)))
rect_shade(xs, y = c(11.5, 12.5), fillcol = cols, linecol = "grey30")
text(xs, rep(10.5, length(xs)), dwgttxt, srt = 90, col = "grey30")
text(mean(xs), 13, "% of observations", col = "grey30")
mtext("B", side = 3, line = -1, cex = 1, adj = 0.05)
dev.off()

### 2017-05-15 paper/main.tex
Added Figure 3 to text and associated text.  Edited Figure 3 to add percetage values for each bin atop whiskers.

%reveals a log-linear relationship wherein resolution strongly constrains extent, such that the ratio between extent and resolution falls as the latter increases (Fig. 2C). As with interval, this pattern indicates that spatial replication diminishes with plot size \cite{kareiva_spatial_1988} in field-based studies, which accounted for 80\% of observations. This stands in contrast to the 6.5\% of observations that were made with remote sensing, which have fine spatial resolutions relative to their large extents. These comprise the majority of observations in the four low density patches (e.g. 0.1-10 ha resolution and 1-10 million ha extent) falling above the denser line that terminates at 100 ha resolution and 1000 ha extent.

A similar tendency for duration to increase with interval was also evident amongst temporally replicated observations (Fig. 2D), where the majority of observations were repeated at daily to decadal intervals and spanned  $\geq$ 1 month to $<$100 years.  The orientation of this concentration shows that observations lasting one year to one decade tend to have corresponding intervals, while those lasting one month to one year have daily to monthly intervals. This suggests that most long-duration observations have just a single temporal replicate. The fainter densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

%A similar linear relationship between sampling interval and temporal duration was not evident (unreplicated observations were excluded because they lacked interval values). For repeated observations, the greatest concentration was among those made at daily to yearly intervals with overall durations of hours to months (Fig. 2D). A second concentration of sub-monthly to yearly observations lasting one second or less is also evident. Very few repeat observations exceeded a year in duration, and these were collected at daily to yearly intervals.

@@

\noindent \textbf{\emph{Insights into the scales of modern ecology}}\\
Despite these caveats, our results provide valuable insight into the spatial and temporal domains being addressed by modern ecological research. Our results show that most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month). Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989}. Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. Taken together, this narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) suggests that the ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}, a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.

>>>>

A similar tendency for duration to increase with interval was also evident amongst temporally replicated observations (Fig. 2D), where the majority of observations were repeated at daily to decadal intervals and spanned  $\geq$ 1 month to $<$100 years.  The orientation of this concentration shows that observations lasting one year to one decade tend to have corresponding intervals, while those lasting one month to one year have daily to monthly intervals. This suggests that most long-duration observations have just a single temporal replicate. The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

To provide further insight into the spatio-temporal domains of ecological observations, we also evaluated the differences in scale between extent and actual extent and between duration and actual duration. To make these comparisons, we calculated the magnitudes of difference (decades) between each dimension and its actual equivalent, and examined how these varied as a function of the scale of the actual measurement (Fig. 3). The majority (80\%) of the assessed observations had actual extents of $\leq$1 ha, which on average was 4 to nearly 8 orders of magnitude smaller than the quantified extent (Fig 3.A).  Actual extent converged with extent above 100,000 ha, but this applies to just 5\% of observations. The actual duration of 65\% of observations is $<$ 1 day, which is on average 3 to nearly 9 orders of magnitude shorter than the time span covered by temporal replicates (Fig. 3B). The two duration measures were effectively the same for the 16\% of observations exceeding one month of actual duration.

\begin{figure}[ht]
\includegraphics[width=1\textwidth]{figures/act_v_eff_diff.png}
\vspace{-0.2 cm}
\caption{The difference between \emph{actual} extent (the summed area of spatial replicates) and extent (A) and \emph{actual} duration (the summed sampling duration across temporal replicates) and duration (B).  Differences are expressed as decades, or how many orders of magnitude greater the extent or duration is then the actual extent or duration, and are summarized (as box plots, with circle in box representing the mean and line the median) in bins representing different levels of actual extent/duration.  The percentages of observations falling within each bin are indicated by the color of the inter-quartile and the numeric value above the upper whisker.}
\label{afoto1}
\end{figure}

@@

\noindent \textbf{\emph{Insights into the scales of modern ecology}}\\
%These results indicate that for most ecological observations are collected with spatially discontinuous replicates that sample very small portions of the area in which they are distributed.

Despite these caveats, our results provide valuable insight into the spatial and temporal domains being addressed by modern ecological research. Our results show that most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month). Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989}. Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. Taken together, this narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) suggests that the ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}, a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.
